Interdependence of a kinase and its cognate substrate plasma membrane nanoscale dynamics underlies Arabidopsis response to viral infection

Version 1

Abstract

Abstract

Plant viruses represent a risk to agricultural production and as only few treatments exist, it is urgent to identify resistance mechanisms and factors. In plant immunity, plasma membrane (PM)-localized proteins are playing an essential role in sensing the extracellular threat presented by bacteria, fungi or herbivores. Viruses being intracellular pathogens, the role of the plant PM in detection and resistance against viruses is often overlooked. We investigated the role of the partially PM-bound Calcium-dependent protein kinase 3 (CPK3) in viral infection and we discovered that it displayed a specific ability to hamper viral propagation over CPK isoforms that are involved in immune response to extracellular pathogens. More and more evidence support that the lateral organization of PM proteins and lipids underlies signal transduction in plants. We showed here that CPK3 diffusion in the PM is reduced upon activation as well as upon viral infection and that such immobilization depended on its substrate, Remorin (REM1.2), a scaffold protein. Furthermore, we discovered that the viral infection induced a CPK3-dependent increase of REM1.2 PM diffusion. Such interdependence was also observable regarding viral propagation. This study unveils a complex relationship between a kinase and its substrate that contrasts with the commonly described co-stabilisation upon activation while it proposes a PM-based mechanism involved in decreased sensitivity to viral infection in plants.

Introduction

Viruses are intracellular pathogens, carrying minimal biological material and strictly relying on their host for replication and propagation. They represent a critical threat to both human health and food security. In particular, potexvirus epidemics like the one caused by pepino mosaic virus dramatically affect crop production1 and the lack of chemical treatments available makes it crucial to develop inventive protective methods. Unlike their animal counterparts, which enter host cells by interacting directly with the plasma membrane (PM), plant viruses have to rely either on mechanical wounding or insect vectors to cross the plant cell wall1. For this reason, only a few PM-localized proteins were identified as taking part in immunity against viruses2,3. Among them, members of the REMORIN (REM) protein family were shown to be involved in viral propagation, with varying mechanisms depending on the studied viral genera4–9. REM proteins are well-known for their heterogeneous distribution at the PM, forming nanodomains (ND), PM nanoscale environments that display a composition different form the surrounding PM10,11. Increasing evidence supports the role of ND in signal transduction, with the underlying idea that the local accumulation of proteins allows amplification and specification of the signal11. For example, Arabidopsis RHO-OF-PLANT 6 accumulates in distinct ND upon osmotic stress and auxin treatment in a dose-dependent way for the latter. Recently, Arabidopsis REM1.2 (later named REM1.2) form clusters upon exposure to the bacterial effector flg22 to support the condensation of Arabidopsis FORMIN 6 and induce actin cytoskeleton remodeling12. However, unlike the canonical mechanism describing the accumulation of proteins in ND upon stimulation12–14, we showed previously that Solanum tuberosum REM1.3 (StREM1.3) ND were disrupted and the diffusion of individual proteins increased in response to a viral infection. StREM1.3 lateral organization in lipid bilayers was also shown to be modified upon its phosphorylation status, both in vitro and in vivo15,16. The role of such protein dispersion upon stimuli is not understood yet and only few similar cases are reported in the literature17,18. REM1.2 was identified as a substrate of the partially membrane-bound CALCIUM-DEPENDENT PROTEIN KINASE 3 (CPK3)19. CPK3 is involved in defense response against herbivores, bacteria and viruses and was recently proposed to be at cross-roads between pattern-triggered immunity and effector-triggered immunity15,20,21. We showed previously that transient over-expression of CPK3 in N. benthamiana was able to hamper potato virus X (PVX, potexvirus) cell-to-cell propagation15. Although partially PM localized, the role of CPK3 PM localization in immunity was never investigated for any pathogen. As the PVX cannot infect Arabidopsis thaliana, the dedicated plant model for genetic studies, we used an alternative virus model able to infect this species, the plantago asiatica mosaic virus (PlAMV)22,23 to investigate the role of CPK3 and of its PM localization in potexvirus propagation. We were able to highlight the specific role of CPK3 among other immune-related CPKs24–26 in this context. Also, we demonstrated that CPK3 membrane localization was required to hamper viral cell-to-cell propagation and, using single-particle tracking photoactivated light microscopy (spt-PALM), we discovered that CPK3 diffusion was reduced upon activation and viral infection. Interestingly, this reduction of PM diffusion depended on the expression of Group 1 REM while viral-induced REM1.2 increased PM diffusion depended on CPK3. Overall, our data allowed us to proposed a model for a PM-localized mechanism involved in the reduction of potexvirus propagation, encouraging to explore the involvement of the PM in viral immunity.

Results

Arabidopsis thaliana calcium-dependent protein kinase 3 (CPK3) specifically restricts PlAMV propagation

We previously showed that transient overexpression of Arabidopsis CPK3 in N. benthamiana leaves restricted the propagation of the potexvirus potato virus X (PVX)15. CPKs are encoded by a large gene family of 34 members in Arabidopsis27. To evaluate the functional redundancy between CPKs regarding viral propagation, a series of Arabidopsis lines independently mutated for CPK1, CPK2, CPK3, CPK5, CPK6 or CPK11, that are involved in plant resistance to various pathogens21,24–26,28, were analyzed in a viral propagation assay. Because PVX does not infect Arabidopsis, we used instead a binary plasmid encoding for the genomic structure of a GFP-tagged PlAMV22,23, a potexvirus that is capable of infecting a wide range of plant hosts, including Arabidopsis. Agrobacterium carrying PlAMV-GFP were infiltrated in A. thaliana leaves and GFP-fluorescent infection foci were observed 5 days post infiltration (dpi) (Figure 1 – figure supplement 1). The following combinations of mutants were tested: the cpk1 cpk2 double mutant25, the cpk5 cpk6 double mutant24, the triple mutant cpk5 cpk6 cpk124 and the two quadruple mutants cpk1 cpk2 cpk5 cpk625 and cpk3 cpk5 cpk6 cpk1126. No significant difference of PlAMV-GFP infection foci area was detected between Col-0, cpk1 cpk2, cpk5 cpk6, cpk1 cpk2 cpk5 cpk6 and cpk5 cpk6 cpk11, demonstrating that CPK1, CPK2, CPK5, CPK6 and CPK11 are not involved in PlAMV propagation (Figure 1A, 1B). However, a 20 % increase of PlAMV-GFP infected area was observed in cpk3 cpk5 cpk6 cpk11 quadruple mutant relative to the control Col-0, which indicates that CPK3 could play a specific role in viral propagation. Since group 1 REMs are known substrates of CPK319, we checked REM1.2 phosphorylation specificity by the CPK isoforms tested in viral propagation (Figure 1 – figure supplement 2). CPK1, 2 and 3 displayed the strongest kinase activity on REM1.2 while CPK5, CPK6 and CPK11 displayed a residual activity. In contrast, all 6 CPKs phosphorylated the generic substrate histone, suggesting some substrate specificity for REM1.2 in vitro. Since CPK1 and CPK2 were described to be mainly localized within the peroxisome29 and endoplasmic reticulum30, respectively, we hypothesize that they likely do not interact in vivo with PM-localized REM1.231.

Arabidopsis thaliana calcium-dependent protein kinase 3 (CPK3) is specifically involved in the restriction of PlAMV cell-to-cell movement.

A. Representative images of PlAMV-GFP infection foci at 5 dpi in the different mutant backgrounds. Scale bar = 500 µm

B. Box plots of the mean area of PlAMV-GFP infection foci 5 days after infection in CPK multiple mutant lines, normalized to the mean area measured in Col-0. Three independent biological repeats were performed, with at least 47 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

C. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-1 and cpk3-2 single mutants and in CPK3 over-expressing lines (Pro35S:CPK3-HA #8.2 and Pro35S:CPK3-HA #16.2), normalized to the mean area measured in Col-0. Three independent biological repeats were performed, with at least 56 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

D. Representative images of A. thaliana plants infected with PlAMV-GFP and imaged with a CCD Camera from 10 to 17 dpi. The region of interest used for measurement of pixel intensity is circled with a white dotted line. Multicolored scale is used to enhance contrast and ranges from blue (low intensity) to red (high intensity). Scale bar = 4 cm.

E. Box plots of the mean cumulated intensity measured in infected leaves in Col-0, cpk3-2 and Pro35S:CPK3-HA #16.2 during systemic viral propagation. Two independent experiments were conducted. Statistical differences could be observed between the genotypes and time-points using a Kruskal-Wallis followed by a Dunn’s multiple comparison test (p<0.05). For clarity, only the results of the statistical test of the comparison of the different time-points (10, 14 and 17 dpi) within a genotype are displayed and are color-coded depending on the genotype.

F. Kinase activity of CPK3 dead variant. Recombinant proteins GST-CPK3 WT and K107M were incubated with REM1.21–118–6His in kinase reaction buffer in the presence of EGTA (−) or 100 µM free Ca2+ (+). Radioactivity is detected on dried gel (upper panel). The protein amount is monitored by Coomassie staining (lower panel).

G. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2 complemented lines cpk3-2/Pro35S:CPK3-myc and cpk3-2/ Pro35S:CPK3K107M-myc. Three independent biological repeats were performed, with at least 51 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001)

To further confirm a role for CPK3 in PlAMV infection, two independent knock-out (KO) lines for CPK3, cpk3-132 and cpk3-219 (Figure 1 – figure supplement 3) along with two independent lines overexpressing CPK3 (i.e., Pro35S:CPK3-HA-OE #8.2 and Pro35S:CPK3-HA-OE#16.224; Figure 1 – figure supplement 3) were infiltrated with PlAMV-GFP. In both cpk3-1 and cpk3-2 KO lines, PlAMV-GFP propagation was significantly enhanced (40 to 60% compared with WT Col-0), whereas the over-expression lines Pro35S:CPK3-HA-OE#8.2 and Pro35S:CPK3-HA-OE#16.224 showed 10% and 20 % restriction of the foci area, respectively (Figure 1C). To assess whether CPK3 regulates viral propagation at the plant level, the propagation of PlAMV-GFP in systemic leaves was assessed in 3-week-old cpk3-2 and Pro35S:CPK3-HA-OE#16.2 lines along with Col-0 at 10, 14 and 17 days post infection (dpi) using a CCD Camera equipped with a GFP filter (Figure 1 – figure supplement 1). We observed that loss of CPK3 led to an increase of PlAMV-GFP propagation in distal leaves during the course of our assay while the overexpression of CPK3 did not hamper PlAMV-GFP to a greater extent than WT Col-0 (Figure 1D and E). This would suggest that CPK3 effect on PlAMV propagation is predominant at the site of infection. For this reason, we’ll focus on foci in local leaves for further study.

CPK3 Lysine 107 functions as ATP binding site and its substitution into methionine abolishes CPK3 activity in vitro21. In good agreement, we observed that CPK3K107M can no longer phosphorylate REM1.2 in vitro (Figure 1F). To test whether CPK3 kinase activity is required for its function during PlAMV infection, we analyzed the propagation of PlAMV-GFP in complementation lines of cpk3-2 with WT CPK3 or with CPK3K107M. While the WT CPK3 fully complemented cpk3-2 mutant, it was not the case with cpk3-2/Pro35S:CPK3K107M, which displayed larger infection foci area compared to WT Col-0 (Figure 1G).

Taken together, these results demonstrate a specific involvement for CPK3 among other previously described immune-related CPKs in limiting PlAMV infection.

PlAMV infection induces a decrease in CPK3 PM diffusion

We next wondered whether CPK3 accumulation is regulated at transcriptional and translational level upon PlAMV infection. RT-qPCR and western blots of CPK3 in Col-0 showed that both transcript and protein levels remained unchanged during PlAMV infection (Figure 2 – figure supplement 1). CPK3 is partially localized within the PM17 and is myristoylated at Glycine 2, a modification required for its association with membranes19. To test whether CPK3 membrane localization is required to hamper PlAMV propagation, we transformed cpk3-2 mutant with either ProUbi10:CPK3-mRFP1.2 or ProUbi10:CPK3G2A-mRFP1.2 (Figure 2A) and tested PlAMV infection. We observed that, in contrary to ProUbi10:CPK3-mRFP1.2, ProUbi10:CPK3G2A-mRFP1.2 did not complement cpk3-2 (Figure 2B). These observations indicate that CPK3 association with the PM is required for its function in inhibiting PlAMV propagation. We next analyzed the organization of CPK3 PM pool in absence or presence of PlAMV-GFP using confocal microscopy. Imaging of the surface of A. thaliana leaf epidermal cells expressing CPK3-mRFP1.2 showed that the protein displayed a heterogeneous pattern at the PM in both conditions (Figure 2C), although the limitation in lateral resolution of confocal microscopy hindered a more detailed analysis of CPK3 PM organization.

CPK3 diffusion decreases upon PlAMV infection in Arabidopsis thaliana

A. Confocal images of the secant view of A. thaliana epidermal cells of the cpk3-2/CPK3-mRFP1.2 or cpk3-2/ProUbi10:CPK3G2A-mRFP1.2. Scale bar = 5µm

B. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2 and complemented lines cpk3-2/ProUbi10:CPK3-mRFP1.2 and cpk3-2/ProUbi10:CPK3G2A-mRFP1.2. Three independent biological repeats were performed, with at least 32 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001).

C. Confocal images of the surface view of A. thaliana epidermal cells of the Col-0/ProUbi10:CPK3-mRFP1.2, infiltrated either with free GFP (“mock”) or PlAMV-GFP. Scale bar = 5µm

D. Representative trajectories of Col-0/CPK3-mEOS3.2 infiltrated either with free GFP (“mock”) or PlAMV-GFP; Scale bar = 2µm

E. Distribution of the diffusion coefficient (D), represented as log(D) for Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 16 cells over the course of three independent experiments.

F. Box plot of the mean peak value extracted from the Gaussian fit of log(D) distribution. Significant difference was revealed using a Mann-Whitney test. *: p<0.05.

G. Mean square displacement (MSD) over time of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Representative trajectories extracted from Figure 2D illustrate each curve. Scale bar = 1µm. Data were acquired from at least 16 cells over the course of three independent experiments.

H. Voronoi tessellation illustration of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. ND are circled in red. Scale bar = 2µm

I. Distribution of the ND diameter of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

J. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 2I. No significant differences were revealed using a Mann-Whitney test.

K. Boxplot of the proportion of Col-0/ProUbi10:CPK3-mEOS3.2 detections found in ND five days after infiltration with free GFP (“mock”) or PlAMV-GFP. No significant differences were revealed using a Mann-Whitney test.

Next, we used single particle tracking phospho-activated localization microscopy (sptPALM) which overcomes the diffraction limit of confocal microscopy and allows to analyze the diffusion and organization of single molecules. We used a translational fusion of CPK3 with the true monomeric photoconvertible fluorescent protein mEOS3.233 expressed in stable transgenic Arabidopsis lines. We imaged these materials in control and upon PlAMV infection. We tracked single molecule trajectories (Figure 2D) from which CPK3 diffusion coefficient (D) was calculated. We observed that CPK3 proteins were overall mobile in control and infected conditions (log(D) > −2; Figure 2E) although CPK3 diffusion was reduced upon PlAMV infection (Figure 2F). Analysis of the mean squared displacement (MSD), describing the surface explored by single molecules overtime, showed that CPK3 displayed a more confined behavior during a PlAMV infection than in healthy conditions (Figure 2G). Additionally, we performed cluster analysis using Voronoï tessellation, a computation method that segments super-resolution images into polygons based on the local molecule density34. Voronoï analysis showed that no difference occurred in CPK3 cluster size or proportion of protein localized in cluster upon viral infection (Figure 2H-K). Taken together, these results show that CPK3 diffusion parameters were modified upon PlAMV infection, although the nano-organization of the proteins was maintained.

CPK3 activation leads to its confinement in PM ND

CPK3 bears an auto-inhibitory domain that folds over the kinase domain and inhibits its kinase activity in the absence of calcium35,36. The truncation of this domain along with the C-terminal regulatory domain results in a calcium-independent, constitutively active CPK3 (CPK3CA)24 that is lethal when stably expressed in Arabidopsis37. For this reason, CPK3CA was transiently expressed in N. benthamiana for further analysis. We observed that although both ProUbi10:CPK3-mRFP1.2 and ProUbi10:CPK3CA-mRFP1.2 were partially cytosolic when transiently expressed in N. benthamiana (Figure 3 – figure supplement 1), ProUbi10:CPK3CA-mRFP1.2 displayed a PM organization in domains discernable by confocal microscopy (Figure 3A). While the mechanism(s) governing the clustering of membrane proteins are not fully described, it is widely accepted that lateral organization involves – to some extent – protein-lipid interactions and lipid-lipid organization10,11,38. We observed that the integrity of ProUbi10:CPK3CA-mRFP1.2 organization relied on sterol and phosphoinositides. Indeed, treatment with fenpropimorph, a well-described inhibitor of sterol biosynthesis39, abolished CPK3CA ND organization (Figure 3 – figure supplement 2), and the co-expression of ProUbi10:CPK3CA-mRFP1.2 with the yeast phosphatidylinositol-4-phosphate (PI4P)-specific phosphatase SAC1 targeted to the PM40, led to sparser and bigger domains (Figure 3 – figure supplement 2), which suggested that PI4P is not required for CPK3CA ND formation but for its regulation.

PlAMV-induced activation of CPK3 in N. benthamiana induces its confinement and clustering in PM domains

A. Confocal images of the surface view of N. benthamiana epidermal cells transiently expressing ProUbi10:CPK3-mRFP1.2, ProUbi10:CPK3-mRFP1.2 + PlAMV-GFP or ProUbi10:CPK3CA-mRFP1.2. Scale bar = 5µm

B. Representative trajectories of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3.2-mEOS3.2 + PlAMV and ProUbi10:CPK3CA-mEOS3.2. Scale bar = 2µm.

C. Distribution of the diffusion coefficient (D), represented as log(D) for ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. Data were acquired from at least 15 cells over the course of three independent experiments.

D. Box plots of the fraction of immobile proteins (log(D)<-2). Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.005).

E. Mean square displacement (MSD) over time of fast (circle) or slow (triangle) diffusing ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP, ProUbi10:CPK3CA-mEOS3.2. Standard error is displayed from three independent experiments.

F. Voronoï tessellation illustration of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. ND are circled in red. Scale bar = 2µm

G. Distribution of the ND diameter of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2

H. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 3G. No significant differences were revealed using a Kruskal-Wallis followed by a Dunn’s multiple comparison test.

I. Boxplot of the proportion of detections found in ND of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. Significant differences were revealed using a Kruskal-Wallis followed by a Dunn’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.005).

We analyzed the dynamics of ProUbi10:CPK3-mEOS3.2 and ProUbi10:CPK3CA-mEOS3.2 by spt-PALM (Figure 3B). We observed that the fraction of immobile ProUbi10:CPK3CA-mEOS3.2 molecules (log(D) < −2) is more abundant than for CPK3-mEOS3.2 (Figure 3C and D). In addition, MSD analysis showed that CPK3CA-mEOS3.2 motion is more confined than CPK3-mEOS3.2 (Figure 3E). Overall, the behavior of CPK3CA-mEOS3.2 is reminiscent of CPK3-mEOS3.2 upon PlAMV infection, suggesting that changes in CPK3 dynamics upon PlAMV infection are linked to its activation. The differences in the amplitude of immobilization of CPK3CA compared to CPK3 upon PlAMV infection may imply that activation of CPK3 upon PlAMV is transient and dynamic. Cluster analysis of CPK3 was performed using tessellation on the localization data obtained with spt-PALM (Figure 3F). During both infected and viral infection conditions, we did not observe any significant differences as compared to CPK3CA (Figure 3G and H). However, CPK3CA displayed a significantly higher number of proteins localized in ND compared to CPK3 (Figure 3I).

All together these observations hint that PlAMV infection leads to the activation of CPK3 and changes its dynamics within the PM.

PlAMV infection induces an increase in REM1.2 PM diffusion

Group 1 REMs are one of the targets of CPK3 and we previously that the restriction of PVX propagation by CPK3 overexpression depended on endogenous group 1 NbREMs15. Four REM isoforms belong to the group 1 in Arabidopsis: REM1.1, REM1.2, REM1.3 and REM1.441. REM1.2 and REM1.3 are amongst the 10% most abundant transcripts in Arabidopsis leaves while REM1.1 was not detected in recently published leaf transcriptomes and proteomes42. Therefore, we focused on the three isoforms REM1.2, REM1.3 and REM1.4. REMs are described as scaffold proteins43, for which physiological function depends on the proteins they interact with and their phosphorylation status44. As recently described in45, REM1.2 and REM1.3 share 95% of their interactome, suggesting that they are functionally redundant. To address this, we isolated single T-DNA mutants rem1.2, rem1.3 and rem1.4 (SALK_117637.50.50.x, SALK_117448.53.95.x and SALK_073841.47.35, respectively) and crossed them to obtain the double mutant rem1.2 rem1.3 and the triple mutant rem1.2 rem1.3 rem1.4 (Figure 4 – figure supplement 1). We did not notice any obvious defects in the growth and development of seedlings and adult plants for the single, double and triple mutants, when grown under our conditions (Figure 4 – figure supplement 2). No difference in PlAMV-GFP propagation could be observed in the single mutants, compared to Col-0 (Figure 4A). However, the double mutant rem1.2 rem1.3 showed a significant increase of infection foci area compared to Col-0, which was further enhanced in rem1.2 rem1.3 rem1.4 triple KO mutant. Such additive effect of multiple mutations shows that REM1.2, REM1.3 and REM1.4 are functionally redundant regarding PlAMV cell-to-cell propagation. Finally, PlAMV-GFP systemic propagation was followed in whole plants every 3-4 days from 10 to 17 dpi and rem1.2 rem1.3 rem1.4 displayed an increased infection surface of systemic leaves compared to Col-0 (Figure 4B and C), suggesting that group 1 REMs are involved in both local and systemic propagation of PlAMV-GFP.

Group 1 REM hampers PlAMV-GFP cell-to-cell propagation and REM1.2 diffusion increases upon infection

A. Box plots of the mean area of PlAMV-GFP infection foci in rem1.2, rem1.3, rem1.4 single mutants along with rem1.2 rem1.3 double mutant and rem1.2 rem1.3 rem1.4 triple mutant. Three independent biological repeats were performed, with at least 36 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

B. Box plots of the mean cumulated intensity measured in infected leaves in Col-0 and rem1.2 rem1.3 rem1.4 during systemic viral propagation. Two independent experiments were conducted. Statistical significance of the difference between Col-0 and rem1.2 rem1.3 rem1.4 at each time point was assessed using a Mann-Whitney test. *: p<0.05, **: p<0.01

C. Representative images of A. thaliana plants infected with PlAMV-GFP and imaged with a CCD Camera from 10 to 17 dpi. Systemic leaves are circled with a white dotted line. Multicolored scale is used to enhance contrast and ranges from blue (low intensity) to red (high intensity). Scale bar = 4 cm.

D. Confocal images of the surface view of A. thaliana epidermal cells of the Col-0/ProUbi10:mRFP1.2-REM1.2, infiltrated either with free GFP (“mock”) or PlAMV-GFP. Scale bar = 5µm

E. Representative trajectories of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm

F. Distribution of the diffusion coefficient (D), represented as log(D) for Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 16 cells over the course of three independent experiments.

G. Box plot of the mean peak value extracted from the Gaussian fit of log(D) distribution. Significant difference was revealed using a Mann-Whitney test. *: p<0.05.

H. Mean square displacement (MSD) over time of Col-0/ProUbi10:mEOS3.2-REM1.2 infiltrated either with free GFP (“mock”) or PlAMV-GFP. Representative trajectories extracted from Figure 4E illustrate each curve. Scale bar=1µm.

I. Voronoi tessellation illustration of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. ND are circled in red. Scale bar = 2µm

J. Distribution of the ND diameter of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

K. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 4J. No significant difference was revealed using a Mann-Whitney test.

L. Boxplot of the proportion of Col-0/ProUbi10:mEOS3.2-REM1.2 detections found in ND five days after infiltration with free GFP (“mock”) or PlAMV-GFP. No significant difference was revealed using a Mann-Whitney test.

Given their role in cell-to-cell viral propagation, we checked whether group 1 REM expression was modified upon infection. RT-qPCR and western blots showed that neither transcripts nor protein levels were modified upon PlAMV infection (Figure 4 – figure supplement 3). Since REM1.2 and REM1.3 share a large part of their interactome and show functional redundancy regarding PlAMV infection, we decided to focus on REM1.2 for further investigations. Confocal imaging of the surface of epidermal cells of Col-0/ProUbi10:mRFP1.2-REM1.2 showed a rather heterogeneous distribution at the PM, although less striking than previously described when observed in root seedlings31 (Figure 4D). REM1.2 was next fused to mEOS3.2 and stably-expressed in Col-0 to conduct spt-PALM (Figure 4E). The D of ProUbi10:mEOS3.2-REM1.2 was significantly increased upon PlAMV infection (Figure 4F and G). Moreover, its MSD was increased to a similar extent as what was previously observed with StREM1.3 during a PVX infection15 (Figure 4H), although REM1.2 is overall more mobile than StREM1.3. Tessellation analysis of protein localization did not show any difference in ND organization, whether in size or regarding the enrichment of proteins in ND (Figure 4I-L).

Taken together, these results show that group 1 REMs are functionally redundant regarding their ability to hamper PlAMV propagation. PlAMV infection promoted an increased diffusion of REM1.2, in the same way as PVX did with StREM1.3. The conservation of such mechanism between plants of different families is indicative of its physiological importance.

PlAMV-induced changes in REM1.2 and CPK3 plasma membrane dynamics are interdependent

Group 1 REMs from A. thaliana were previously identified as in vitro substrates of CPK317. Moreover, untargeted immunoprecipitation experiments coupled to mass spectrometry identified CPK3 as an interactor of REM1.2 in A. thaliana45. We wanted to assess the functional link between CPK3 and group 1 REM in potexvirus propagation by knocking out CPK3 into the rem1.2 rem1.3 rem1.4 mutant background. We isolated two independent CRISPR-generated rem1.2 rem1.3 rem1.4 cpk3 #1 and rem1.2 rem1.3 rem1.4 cpk3 #2 quadruple mutants (Figure 5 – figure supplement 1). We did not observe any developmental defect in these lines when grown under controlled conditions (Figure 5 – figure supplement 2). The analysis of PlAMV-GFP propagation showed that no significant additive effect could be observed between the quadruple mutant lines, cpk3-2 and rem1.2 rem1.3 rem1.4 (Figure 5A and 5B). This indicates that group 1 REMs and CPK3 function in the same signaling pathway to inhibit PlAMV propagation.

Group 1 REMs and CPK3 are in the same functional pathway and regulate each other PM diffusion upon PlAMV infection

A. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2, rem1.2 rem1.3 rem1.4 triple mutant and rem1.2 rem1.3 rem1.4/cpk3 #1 and #2 quadruple mutants. Three independent biological repeats were performed, with at least 23 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001).

B. Representative images of PlAMV-GFP infection foci at 5 dpi in the different mutant backgrounds. Scale bar = 500µm.

C. Representative trajectories of cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with either free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm.

D. Distribution of the diffusion coefficient (D), represented as log(D) for cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with either free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 11 cells over the course of three independent experiments.

E. Box plot representing the mean peak value extracted from the Gaussian fit of the distribution of the diffusion coefficient (D) represented in Figure 5D. No significant difference was revealed using a Mann-Whitney test.

F. Mean square displacement (MSD) over time of cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

G. Representative trajectories of rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm.

H. Distribution of the diffusion coefficient (D), represented as log(D) for rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 10 cells over the course of three independent experiments.

I. Box plot representing the mean peak value extracted from the Gaussian fit of the distribution of the diffusion coefficient (D) represented in Figure 5H. No significant difference was revealed using a Mann-Whitney test.

J. Mean square displacement (MSD) over time of rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

We wanted to know whether the increased diffusion of REM1.2 observed on PlAMV infection was dependent on CPK3. Using spt-PALM, we obtained the diffusion parameters of ProUbi10:REM1.2-mEOS3.2 expressed in the cpk3-2 mutant background. Strikingly, we observed that both the D and MSD of REM1.2 were not anymore affected during a viral infection (Figure 5C-F), showing that REM1.2 PM lateral diffusion upon PlAMV infection depends on CPK3. In a similar manner as in Col-0, REM1.2 clustering upon PlAMV infection in cpk3-2 background did not display any difference to the mock-infected plants (Figure 5 – figure supplement 3).

Moreover, we wondered whether the reciprocal effect was true for the diffusion of CPK3 in the absence of group 1 REMs. Similarly, the D and the MSD of mEOS3.2-CPK3 in rem1.2 rem1.3 rem1.4 triple KO background remained the same during an infection compared to control condition (Figure 5G-J), unlike what was observed in a Col-0 background (Figure 2D-G). This result indicated that the confinement of CPK3 proteins upon viral infection depended on the presence of group 1 REMs. However, contrarily to the Col-0 background, the rem1.2 rem1.3 rem1.4 displayed smaller CPK3 ND size with reduced protein concentration upon PlAMV infection (Figure 5 – figure supplement 3). This showed that group 1 REMs might play a role in CPK3 domain stabilization upon viral infection. We investigated whether CPK3 and REM would colocalize in absence or presence of the virus. Using confocal microscopy, we showed that they randomly colocalized in both situations (Figure 5 – figure supplement 4), the interaction between the kinase and its substrate probably occurring in a narrow spatiotemporal window.

Taken all together, those results show a strong inter-dependence of group 1 REMs and CPK3 both in their physiological function and in their PM lateral diffusion upon PlAMV infection.

Discussion

CPK3 specific role in viral immunity is supported by its PM organization

Although calcium-mediated signaling is suspected to be involved in viral immunity, only few calcium-modulated proteins are described to play a role in viral propagation15,46. We showed here the crucial role of CPK3 over other immunity-related CPK isoforms24,25 by a reverse genetic approach in Arabidopsis. The precise role of CPK3 in viral immunity remains to be determined. CPK3 phosphorylates actin depolymerization factors to modulate the actin cytoskeleton21,47, a key player in host-pathogen interaction48. In particular, potexviruses induce the remodeling of the actin cytoskeleton to organize the key steps of their cycle, whether it is replication, intra-cellular movement or cell-to-cell propagation49,50. PlAMV replication and/or movement could be affected by CPK3-mediated alteration of the cytoskeleton mesh.

The specificity of a calcium-dependent kinase in a given biological process is determined by its expression pattern, subcellular localization and substrate specificity51. Controlled subcellular localization ensures proximity with either stimulus or substrate and here we showed that, similarly to Arabidospis CPK652 and Solanum tuberosum CPK553, the disruption of CPK3 membrane anchorage led to a loss-of-function phenotype (Figure 2B). Interestingly, we observed a reduction in CPK3 PM diffusion upon PlAMV infection, suggesting that not only membrane localization but also protein organization at the PM is important during viral immunity. Moreover, PlAMV-induced CPK3 PM confinement was reminiscent of the diffusion parameters displayed by CPK3CA, hinting that viral infection, kinase activation and lateral diffusion are linked. However, it is necessary to remain careful as a truncated protein deprived of its auto-inhibitory domain does not reflect the controlled and context-dependent activation of CPK3. Indeed, stable expression of CPK3CA was previously shown to be lethal37. CPK3CA ever-activated state might lead to stable or unspecific interaction with protein partners along with erratic phosphorylation of substrates, which could explain the PM domains formed by CPK3CA at the PM.

CPK3 nanoscale dynamics upon viral infection might offer another layer of specificity to convey the appropriate response to a given stimulus by ensuring proximity with specific regulators or substrates. It would be interesting to explore whether the reduction of CPK3 diffusion observed upon PlAMV infection is specific to this virus or if it can be extended to other viral species, genera and even pathogens. Indeed, it was recently shown that CPK3 transcription enhanced upon infection by viruses from different genera54. Moreover, CPK3 regulates herbivore responses by phosphorylating transcription factors20, is activated by flg22 in protoplasts19 and is proposed to be the target of a bacterial effector to disrupt immune response21. Finally, the diffusion and clustering of other PM-localized CPKs could be investigated as no experimental data exist yet regarding their PM nano-organization. It would be especially relevant to describe these parameters for the CPK isoforms phosphorylating ND-organized NADPH oxidases14,25,28.

Potential functions for REM1.2 increased diffusion upon PlAMV

REM proteins display a wide range of physiological functions and are proposed to function as scaffold proteins43,44,55. Group 1 REMs have been shown to be involved in viral immunity, playing apparent contradictory roles depending on the virus genera8,44,56. Herein, we show through a combination of genetic and biochemical analysis that the three most expressed isoforms of group 1 Arabidopsis REMs were functionally redundant in inhibiting PlAMV propagation (Figure 4A). REMs are anchored at the PM through their C-terminal sequence44,56–58 but despite display a similar PM attachment, potato StREM1.3 is static and form well defined membrane compartments57 while Arabidopsis REM1.2 appeared mobile in leaves, with small and potentially short-lived membrane domains of around 70nm (Figure 4F-J). Beyond clear organizational differences, both REM1.2 and StREM1.3 showed an increased mobility upon viral infection in N. benthamiana and Arabidopsis15 (Figure 4C-F), which is at contrast with the canonical model linking protein activation with its stabilization into ND11,59. Particularly, REM1.2 was recently shown to form ND upon elicitation by a bacterial effector12 or upon exposition to bacterial membrane structures60. Conservation across plant and virus species of such increased PM diffusion of group 1 REMs indicates that this specific mechanism is crucial for plant response to potexviruses, although its role remains to be deciphered. It was suggested, in the context of tobacco rattle virus infection, that REM1.2 clusters led to an increased lipid order of the PM and a morphological modification of plasmodesmata (PD), inducing a decrease of PD permeability31. Moreover, we showed that StREM1.3 modulated the formation of lipid phases in vitro61 while Medicago truncatula SYMREM1 was recently shown to stabilize membrane topology changes in protoplasts62. Therefore, REM1.2 increased diffusion upon viral infection might alter lipid organization, with consequences on PM and PD-localized proteins63. Based on REM’s putative scaffolding role43,64, its increased PM diffusion might modify the stability of REM-supported complexes and induce subsequent signaling, similarly to the way Oryza sativa REM4.1 orchestrates the balance between abscissic acid and brassinosteroids pathways by interacting with different protein kinases65.

A PM-localized mechanism involved in viral propagation hampering

Increasing evidence supports the role of PM proteins in plant antiviral mechanism66. However, while the nano-organization of PM proteins involved in bacterial or fungal immunity begins to be addressed12,60,67,68, there is still only scarce information in a viral infection context. Our observations pointed towards an interdependence of REM and CPK3 in their PM diffusion upon PlAMV infection. In particular, CPK3-dependent REM increased diffusion in a potexvirus infection is consistent with our previously reported PM diffusion increase of StREM1.3 phosphomimetic mutant, reminiscent of StREM1.3 behavior upon viral infection15. Moreover, we recently published that in vitro CPK3-phosphorylated StREM1.3 presented disrupted domains on model membranes compared to the non-phosphorylated protein16. The data presented in this paper further support the predominant role of CPK3 in viral-induced REM1.2 diffusion (Figure 5C-F). Since the actin cytoskeleton was shown to favor nanometric scale ND including those of group 1 REM69, CPK3 mediated regulation of the actin cytoskeleton could regulate REM1.2 PM nanoscale organization. We also discovered the essential role of group 1 REMs in the lateral organization of CPK3 upon viral infection (Figure 5G-J), which is further supported by the difference between CPK3 clustering parameters when expressed in N. benthamiana or in Arabidopsis: CPK3 was enriched in ND upon PlAMV infection when transiently expressed in N. benthamiana but not in Arabidopsis (Figure 2H, 2K, 3F and 3I). The discrepancy between both species could be linked to N. benthamiana REM1 ability to form stable ND, discernable by confocal microscopy5 while we observed small and unstable domains of REM1.2 in Arabidopsis leaves (Figure 4I-K). Moreover, CPK3 ND organization was disrupted in rem1.2 rem1.3 rem1.4 triple KO background upon PlAMV infection (Figure 5 – figure supplement 4), indicating that group1 REMs are crucial for the spatial organization of CPK3 during a viral infection. Furthermore, the sensitivity of CPK3CA domains to the same lipid inhibitors as StREM1.357 (Figure 3 – figure supplement 2) hints that REM1.2 lipid environment might be required for CPK3 nanoscale organization.

Although the data obtained in this paper does not allow to determine the precise sequence of events orchestrating REM1.2 and CPK3 dynamic interaction, hypotheses can be formulated (Figure 6). REM and CPK3 interact in the absence of any stimulus15,45 but as they display drastically different diffusion parameters (Figure 2E, Figure 4F), likely only a small part of proteins interact at basal state. Upon viral infection, CPK3 activation leads to its REM-dependent confinement (Figure 2 and Figure 5). Whether CPK3 activation is a cause, a consequence or is concomitant to its confinement cannot be determined here. When activated, CPK3 phosphorylates its substrates, including REM1.2, which would then be more mobile, as stated before. In addition to the above-mentioned putative roles of REM’s increased diffusion, such “kiss-and-go” mechanism between a kinase and its substrate might also be considered as a negative feedback to hamper constitutive activation of the system. It was recently shown that the PM organization of a receptor-like kinase and its co-receptor relied on the expression of a scaffold protein68,70, REM might go beyond the role of substrate and be essential to maintain the required PM environment of its cognate kinase to ensure proper signal transduction. Although the complexity of potexvirus’ lifecycle as well as technical limitations do not allow to go beyond this model yet, our data supports the significance of fine PM compartmentalization and spatio-temporal dynamics in signal transduction upon viral infection in plants while it explores new concepts underlying plant kinases PM organization.

Proposed “kiss-and-run” model describing REM1.2 and CPK3 interdependence

Upper panel: in healthy conditions, REM1.2 displays a smaller mobility at the PM than CPK3. Random interactions between the two proteins can occur.

Middle panel: upon PlAMV infection, CPK3 is activated. CPK3 activation and confinement are related but whether activation of the protein occurs before or after its confinement is not deciphered yet. Activation results in phosphorylation (+P) of local substrates, including REM.

Lower panel: REM phosphorylation induces its diffusion at the PM. The period during which CPK3 remains confined is not determined yet.

Supporting information

eLife Assessment

This important study describes how the composition and stabilization of nanodomains in the plasma membrane are an integral part of plant defence against viruses, with a focus on the calcium-dependent kinase CPK3 and its apparent interaction with a plasma-membrane nano domain scaffold protein from the remorin family. While the evidence for a specific role of CPK3 in limiting viral spread is convincing, the claims regarding the CPK3-remorin interaction would be strengthened by additional experimental support. The work, which will be of interest to plant cell biologists and plant virology, opens new avenues for understanding the role of plasma membrane nanodomains in limiting viral spread.

Reviewer #1 (Public Review):

Summary:

How plants perceive their environment and signal during growth and development is of fundamental importance for plant biology. Over the last few decades, nano domain organisation of proteins localised within the plasma-membrane has emerged as a way of organising proteins involved in signal pathways. Here, the authors addressed how a non-surface localised signal (viral infection) was resisted by PM localised signalling proteins and the effect of nano domain organisation during this process. This is valuable work as it describes how an intracellular process affects signalling at the PM where most previous work has focused on the other way round, PM signalling effecting downstream responses in the plant. They identify CPK3 as a specific calcium dependent protein kinase which is important for inhibiting viral spread. The authors then go on to show that CPK3 diffusion in the membrane is reduced after viral infection and study the interaction between CPK3 and the remorins, which are a group of scaffold proteins important in nano domain organisation. The authors conclude that there is an interdependence between CPK3 and remorins to control their dynamics during viral infection in plants.

Strengths:

The dissection of which CPK was involved in the viral propagation was masterful and very conclusive. Identifying CPK3 through knockout time course monitoring of viral movement was very convincing. The inclusion of overexpression, constitutively active and point mutation non functioning lines further added to that.

Weaknesses:

My main concerns with the work are twofold.

1. Firstly, the imaging described and shown is not sufficient to support the claims made. The PM localisation and its non-PM localised form look similar and with no PM stain or marker construct used to support this. The sptPALM data conclusions are nice and fit the narrative. However, no raw data or movie is shown, only representative tracks. Therefore the data quality cannot be verified and in addition, the reporting of number of single particle events visualised per experiment is absent, only number of cells imaged is reported. Therefore it is impossible for the reader to appreciate the number of single molecule behaviours obtained and hence the quality of the data.

2. Secondly, remorins are involved in a lot of nano domain controlled processes at the PM. The authors have not conclusively demonstrated that during viral infection the remorin effects seen are solely due to its interaction with CPK3. The sptPALM imaging of REM1.2 in a cpk3 knockout line goes part way to solve this but more evidence would strengthen it in my opinion. How do we not know that during viral infection the entire PM protein dynamics and organisation are altered? Or that CPK3 and REM are at very distant ends of a signalling cascade. Negative control experiments are required here utilising other PM localised proteins which have no role during viral infection. In addition, if the interaction is specific, the transiently expressed CPK3-CA construct (shown to from nano domains) should be expressed with REM1.2-mEOS to show the alterations in single particle behaviour occur due to specific activations of CPK3 and REM1.2 in the absence of PIAMV viral infection and it is not an artefact of whole PM changes in dynamics during viral infection.

In addition, displaying more information throughout the manuscript (such as raw particle tracking movies and numbers of tracks measured) on the already generated data would strengthen the manuscript further.

Overall, I think this work has the potential to be a very strong manuscript but additional reporting of methods and data are required and additional lines of evidence supporting interaction claims would significantly strengthen the work and make it exceptional.

Reviewer #2 (Public Review):

Summary:

The paper provides evidence that CPK3 plays a role in plant virus infection, and reports that viral infection is accompanied by changes in the dynamics of CPK3 and REM1.2, the phosphorylation substrate of CPK3, in the plasma membrane. In addition, the dynamics of the two proteins in the PM are shown to be interdependent.

Strengths:

The paper contains novel, important information.

Weaknesses:

The interpretation of some experimental data is not justified, and the proposed model is not fully based on the available data.

Reviewer #3 (Public Review):

Summary:

This study examined the role that the activation and plasma membrane localisation of a calcium dependent protein kinase (CPK3) plays in plant defence against viruses.

The authors clearly demonstrate that the ability to hamper the cell-to-cell spread of the virus P1AMV is not common to other CPKs which have roles in defence against different types of pathogens, but appears to be specific to CPK3 in Arabidopsis. Further they show that lateral diffusion of CPK3 in the plasma membrane is reduced upon P1AMV infection, with CPK3 likely present in nano-domains. This stabilisation however, depends on one of its phosphorylation substrates a Remorin scaffold protein REM1-2. However, when REM1-2 lateral diffusion was tracked, it showed an increase in movement in response to P1AMV infection. These contrary responses to P1AMV infection were further demonstrated to be interdependent, which led the authors to propose a model in which activated CPK3 is stabilised in nano-domains in part by its interaction with REM1.2, which it binds and phosphorylates, allowing REM1-2 to diffuse more dynamically within the membrane.

The likely impact of this work is that it will lead to closer examination of the formation of nano-domains in the plasma membrane and dissection of their role in immunity to viruses, as well as further investigation into the specific mechanisms by which CPK3 and REM1-2 inhibit the cell-to-cell spread of viruses.

Strengths:

The paper provided compelling evidence about the roles of CPK3 and REM1-2 through a combination of logical reverse genetics experiments and advanced microscopy techniques, particularly in single particle tracking.

Weaknesses:

There is a lack of evidence for the downstream pathways, specifically whether the role that CPK3 has in cytoskeletal organisation may play a role in the plant's defence against viral propagation. Also, there is limited discussion about the localisation of the nano-domains and whether there is any overlap with plasmodesmata, which as plant viruses utilise PD to move from cell to cell seems an obvious avenue to investigate.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

How plants perceive their environment and signal during growth and development is of fundamental importance for plant biology. Over the last few decades, nano domain organisation of proteins localised within the plasma-membrane has emerged as a way of organising proteins involved in signal pathways. Here, the authors addressed how a non-surface localised signal (viral infection) was resisted by PM localised signalling proteins and the effect of nano domain organisation during this process. This is valuable work as it describes how an intracellular process affects signalling at the PM where most previous work has focused on the other way round, PM signalling effecting downstream responses in the plant. They identify CPK3 as a specific calcium dependent protein kinase which is important for inhibiting viral spread. The authors then go on to show that CPK3 diffusion in the membrane is reduced after viral infection and study the interaction between CPK3 and the remorins, which are a group of scaffold proteins important in nano domain organisation. The authors conclude that there is an interdependence between CPK3 and remorins to control their dynamics during viral infection in plants.

Strengths:

The dissection of which CPK was involved in the viral propagation was masterful and very conclusive. Identifying CPK3 through knockout time course monitoring of viral movement was very convincing. The inclusion of overexpression, constitutively active and point mutation non functioning lines further added to that.

Weaknesses:

My main concerns with the work are twofold.

(1) Firstly, the imaging described and shown is not sufficient to support the claims made. The PM localisation and its non-PM localised form look similar and with no PM stain or marker construct used to support this. The sptPALM data conclusions are nice and fit the narrative. However, no raw data or movie is shown, only representative tracks. Therefore, the data quality cannot be verified and in addition, the reporting of number of single particle events visualised per experiment is absent, only number of cells imaged is reported. Therefore, it is impossible for the reader to appreciate the number of single molecule behaviours obtained and hence the quality of the data.

(2) Secondly, remorins are involved in a lot of nanodomain controlled processes at the PM. The authors have not conclusively demonstrated that during viral infection the remorin effects seen are solely due to its interaction with CPK3. The sptPALM imaging of REM1.2 in a cpk3 knockout line goes part way to solve this but more evidence would strengthen it in my opinion. How do we not know that during viral infection the entire PM protein dynamics and organisation are altered? Or that CPK3 and REM are at very distant ends of a signalling cascade. Negative control experiments are required here utilising other PM localised proteins which have no role during viral infection. In addition, if the interaction is specific, the transiently expressed CPK3-CA construct (shown to from nano domains) should be expressed with REM1.2-mEOS to show the alterations in single particle behaviour occur due to specific activations of CPK3 and REM1.2 in the absence of PIAMV viral infection and it is not an artefact of whole PM changes in dynamics during viral infection.

In addition, displaying more information throughout the manuscript (such as raw particle tracking movies and numbers of tracks measured) on the already generated data would strengthen the manuscript further.

Overall, I think this work has the potential to be a very strong manuscript but additional reporting of methods and data are required and additional lines of evidence supporting interaction claims would significantly strengthen the work and make it exceptional.

Reviewer #2 (Public Review):

Summary:

The paper provides evidence that CPK3 plays a role in plant virus infection, and reports that viral infection is accompanied by changes in the dynamics of CPK3 and REM1.2, the phosphorylation substrate of CPK3, in the plasma membrane. In addition, the dynamics of the two proteins in the PM are shown to be interdependent.

Strengths:

The paper contains novel, important information.

Weaknesses:

The interpretation of some experimental data is not justified, and the proposed model is not fully based on the available data.

Reviewer #3 (Public Review):

Summary:

This study examined the role that the activation and plasma membrane localisation of a calcium dependent protein kinase (CPK3) plays in plant defence against viruses.

The authors clearly demonstrate that the ability to hamper the cell-to-cell spread of the virus P1AMV is not common to other CPKs which have roles in defence against different types of pathogens, but appears to be specific to CPK3 in Arabidopsis. Further they show that lateral diffusion of CPK3 in the plasma membrane is reduced upon P1AMV infection, with CPK3 likely present in nano-domains. This stabilisation however, depends on one of its phosphorylation substrates a Remorin scaffold protein REM1-2. However, when REM1-2 lateral diffusion was tracked, it showed an increase in movement in response to P1AMV infection. These contrary responses to P1AMV infection were further demonstrated to be interdependent, which led the authors to propose a model in which activated CPK3 is stabilised in nano-domains in part by its interaction with REM1.2, which it binds and phosphorylates, allowing REM1-2 to diffuse more dynamically within the membrane.

The likely impact of this work is that it will lead to closer examination of the formation of nano-domains in the plasma membrane and dissection of their role in immunity to viruses, as well as further investigation into the specific mechanisms by which CPK3 and REM1-2 inhibit the cell-to-cell spread of viruses.

Strengths:

The paper provided compelling evidence about the roles of CPK3 and REM1-2 through a combination of logical reverse genetics experiments and advanced microscopy techniques, particularly in single particle tracking.

Weaknesses:

There is a lack of evidence for the downstream pathways, specifically whether the role that CPK3 has in cytoskeletal organisation may play a role in the plant's defence against viral propagation. Also, there is limited discussion about the localisation of the nano-domains and whether there is any overlap with plasmodesmata, which as plant viruses utilise PD to move from cell to cell seems an obvious avenue to investigate.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Viral spread work in CPK mutants with time courses is beautiful!

Regarding my public points on my issues with the imaging:

- Figure 2A shows 'PM' localisation of CPK3 and 'non-PM' imaging of CPK3-G2A. The images are near identical both showing cell outlines and cytoplasmic strands. Here a PM marker (such as Lti6B) tagged with a different fluorophore or PM stain should be used in conjunction with surface views (such as in Figure 2C) to show it really is at the PM and the G2A line is not.

Impaired membrane localization of CPK3-G2A is documented in Mehlmer et al., 2010 using microsomal fractionation. Although Figure 2A main purpose is to show correct expression of the constructs in the lines used for PlAMV propagation (Figure 2B), we replaced the images with wider view pictures to be more representative of the subcellular localization of CPK3 and CPK3-G2A.

- Regarding Figure 2C, this is extremely noisy and PM heterogeneity is barely observable over the noise from the system (looking at the edges of surface imaged). You mention low resolution was an issue. I notice from the methods you have taken confocal images on an Zeiss 880 with Airyscan. These images must be confocal but If imaged with Airyscan the PM heterogeneity would be much clearer (see work from John Runions lab).

Indeed, these are tangential views images obtained by Zeiss 880 with Airyscan. Based on tessellation analysis (Figure 2H-J), CPK3 is rather homogeneously distributed and forms ND of around 70nm of diameter. Objects of such size cannot be resolved using pixel reassignment methods such as Airyscan. Note also that AtREM in our study are less heterogeneously distributed than what was described in the literature for StREM1.3.

- Regarding all sptPALM data. At least an example real data image and video is required otherwise the data can’t be assessed. The work of Alex Martiniere (sptPALM) or Alex Jonson (TIRF) all show raw data so the reader can appreciate the quality of the data. In addition, number of events (particles tracked) has to be shown in the figure legend, not just number of cells otherwise was one track performed per cell? Or 10,000? Obviously where the N sits in this range gives the reader more or less confidence of the data.

We agree and we added example videos of sptPALM experiments in the supplementary data, we also indicated the number of tracked particles in the figure legends.

- On a slight technical aside, how do you know the cells being imaged for sptPALM with PIAMV are actually infected with the virus? In Fig 2C you use a GFP tagged version but in sptPALM you use none tagged. I think a sentence in methods on this would help clarify.

PlAMV-GFP was used for spt-PALM experiment and cell infection was assessed during PALM experiment. This is now precised in the corresponding figures and methods.

- I also have a concern over some of the representative images showing the same things between different figures. Your clustering data in 3F looks very convincing. However, in Figure 2H the mock and PIAMV-GFP look very similar. How is Figure 3F so different for the same experiment? Especially considering the scale bars are the same for both figures. Same for CPK3-mRFP1.2 in Fig 2C and 3A, the same thing is being imaged, at the same scale (scale bars same size) but the images are extremely different.

Figure 2 data were generated using CPK3 stably expressed in A. thaliana while Figure 3 data were obtained upon transient over-expression of CPK3 in N. benthamiana. We do not have a clear explanation for such a difference in CPK3 PM behavior, it could lie on a different PM composition or actin organization between those two species, this point is now addressed in the discussion.

- Line 193&194 - you state that the CA CPK3 is reminiscent of the CPK3 upon PIAMV expression. I don't agree, while CPK3CA is less mobile (2D), the MSD shows it is in-between CPK3 and CPK3 + PIAMV. Therefore, can’t the opposite also be true? That overall the behaviour of CPK3-CA is reminiscent of WT CPK. I think this needs rewording.

We agree and we reworded that part

- Line 651 - what numerical aperture are you using for the lens during confocal microscopy. This is fundamentally important information directly related to the reproducibility of the work. You report it for the sptPALM.

The numerical aperture is now indicated in the methods.

Regarding my bigger point about specific interactions between CPK3 and remorin during viral infection to strengthen your claim the following need doing. I am not suggesting you do all of these but at least two would significantly enhance the paper.

(1) Image a none related PM protein during viral infection using sptPALM and demonstrate that its behaviour is not altered (such as lti6b). This would show the affects on remorin behaviour are specific to CPK3 and not a whole scale PM alteration in dynamics due to viral infection.

(2) Two colour SPT imaging of CPK3 and REM1.2. You show in absence of proteins (knockouts effect on each other) but your only interaction data is from a kinase assay (where CPK1 and 2 also interact, even though they are not localised at the same place) and colocalisation data (see below). A two colour SPT imaging experiment showing interaction and clustering of CPK3 and REM1.2 with each other and the change in their behaviours when viral infected and simultaneously imaged would address all of my concerns.

- On another note, the co-localisation data (fig 5 sup 4) needs additional analysis. I would expect most PM proteins to show the results you show as the data is very noisy. In order to improve I would zoom in to fill the field of view and then determine correlation and also when one image is rotated 90 degrees (as described in Jarsch et al., plant cell) to enhance this work.

(3) In the absence of viral infection, but presence of CPK3-CA, is sptPALM REM1.2 behaviour in the PM altered, if so then the interaction is specific and changes in remorin dynamics are not due to whole scale PM changes during viral infection and the manuscript substantially strengthened.

(4) Building on from 3), if you have a CPK3 mutated with both CPK3-CA and G2A this would be constitutively active and non-PM localised and as such should not affect Remorin behaviour if your model is true, this would strengthen the case significantly but I appreciate is highly artificial and would need to be done transiently.

Regarding the first point, since the role of PM proteins involved in potexvirus infection is barely assessed, picking a non-related PM protein might be tricky. The data obtained with mEOS3.2-REM1.2 expressed in cpk3 null-mutant point towards a specific role of CPK3 in PlAMV-induced REM1.2 diffusion and not a general alteration of PM protein behavior.

Regarding the second point, we already reported the in vivo interaction between AtCPK3CA and AtREM1.2/AtREM1.3 by BiFC in N.benthamiana (Perraki et al 2018) and AtCPK3 was shown to co-IP with AtREM1.2 (Abel et al, 2021). While we agree on the relevance of doing dual color sptPALM with CPK3 and REM1.2, it is so far technically challenging and we would not be able to implement this in a timely manner. For the colocalization, although the whole cell is displayed in the figure, the analysis was performed on ROI to fill the field of analysis.

We agree with the relevance of adding the colocalization analysis of randomized images (mTagBFP2 channel rotated 90 degrees), this is now added to Figure 5 – supplement figure 5.

Finally, for the third and fourth points, spt-PALM analysis of REM1.2 in presence of CPK3-CA and CPK3-CA-G2A was performed (Figure 5 - figure supplement 4). The results suggest a specific role of CPK3-CA in REM1.2 diffusion.

Minor points:

Line 59 - from, I think you mean from.

Line 63 - Reference needed after latter.

Line 68 - Reference required after viral infection.

Line 85 - Propose not proposed.

Line 156 - Allowed us to not allows to.

Line 204 - add we previously 'demonstrated'

Line 622 and 623 - You say lines obtained from Thomas Ott. This is very odd phrasing considering he is an author. I appreciate citing the work producing the lines but maybe reword this

These points were corrected, thank you.

Reviewer #2 (Recommendations For The Authors):

The paper provides evidence that CPK3 plays a role in plant virus infection, and reports that viral infection is accompanied by changes in the dynamics of CPK3 and REM1.2, the phosphorylation substrate of CPK3, in the plasma membrane. In addition, the dynamics of the two proteins in the PM are shown to be interdependent. The paper contains novel, important information that can undoubtedly be published in eLife. However, I have some concerns that should be addressed before it can be accepted for publication.

Major concerns

When the authors say that CPK3 plays a role in viral propagation, it should be clarified what is meant by 'propagation', - replication of the viral genome, its cell-to-cell transport, or long-distance transport via the phloem. By default the readers will tend to assume the former meaning. In my opinion, the term 'propagation' is misleading and should be avoided.

We purposely chose the term “propagation” because it sums replication and cell-to-cell movement. Nevertheless, we previously showed that group 1 StREM1.3 doesn’t alter PVX replication (Raffaele et al., 2009 The Plant Cell). In this paper, as we do not investigate the role of AtREM1.2 or AtCPK3 in the replication of the viral PlAMV genome, we cannot state that these proteins are strictly involved in cell-to-cell movement of the virus.

The authors show that viral infection is associated with decreased diffusion of CPK3 and increased diffusion of REM1.2 in the PM. However, it remains unclear whether these changes are related to partial resistance to viral infection involving CPK3 and REM1.2, or whether they are simply a consequence of viral infection that may lead to altered PM properties and altered dynamics of PM-associated proteins. Therefore, the model presented in Fig. 6 appears to be entirely speculative, as it postulates that changes in CPK3 and REM1.2 dynamics are the cause of suppressed virus 'propagation'. In addition, the model implies that a decrease in CPK3 mobility leads to activation of its kinase activity. This view is not supported by experimental data (see my next comment). The model should be deleted (both as the figure and its description in the Discussion) or substantially reworked so that it finally relies on existing data.

For the first point, the results obtained from the additional experiments proposed by reviewer #1 supports the hypothesis of a direct impact of CPK3 on REM1.2 diffusion (Figure 5 - figure supplement 4).

We agree with the second point and reworked the model to remove the link between CPK3 activation and its increased diffusion.

The statement that 'changes in CPK3 dynamics upon PlAMV infection are linked to its activation' (line 194) is based on a flawed logic, and the conclusion in this section of Results ('changes in CPK3 dynamics upon PlAMV infection are linked to its activation') is incorrect, as it is not supported by experimental data. In fact, the authors show that CPK3 dynamics and clustering upon viral infection is somewhat reminiscent of the behavior of a CPK3 deletion mutant, which is a constitutively active protein kinase. However, this partial similarity cannot be taken as evidence that CPK3 dynamics upon PlAMV infection are related to its activation. Furthermore, the authors emphasize the similarity of the mutant and CPK3 in infected cells without taking into account a drastic difference in their localization (Fig. 3A, middle and right panels) showing that the reduced dynamics or the compared proteins may have different causes. I suggest the removal of the section 'CPK3 activation leads to its confinement in PM ND' from the paper, as the results included in this section are not directly related to other data presented.

The PM lateral organization of PM-bound CPKs in their native or constitutively active form as well as the role of lipid in such phenomenon was never shown before. We believe that this section contains relevant information for the community. We kept the section but reworded it to tamper the correlation made between CPK3 PM organization upon viral infection and its activation.

Line 270 - 'group 1 REMs might play a role in CPK3 domain stabilization upon viral infection'. This is an overstatement. The size of the CPK3-containing NDs may have no correlation with their stability.

We reworded the sentence.

Minor points

Line 204 - we previously that Line 234 and hereafter - "the D" sounds strange. Suggest using "the diffusion coefficient".

This was reworded.

Reviewer #3 (Recommendations For The Authors):

The authors have previously demonstrated that there was an increase in REM1.2 localisation to plasmodesmata under viral challenge. It would be useful to see if there was any co-localisation of REM1.2 and CPK3 with plasmodesmata in response to PlAMV and how this is affected in the mutants. This could be carried out relatively simply using aniline blue.

These experiments were added to the supplementary data of Figure 2 – figure supplement 2. and Figure 4 – figure supplement 4. , no enrichment of CPK3 or REM1.2 at plasmodesmata could be observed upon PlAMV infection.

Fig 3 supplementary figure 2 would be better incorporated into the main body of Figure 3 as this underpins discussion on the involvement of lipids such as sterols in the formation of nanodomains.

We moved Figure 3 – Supplementary figure 2 to the main body of Figure 3.

Minor corrections:

Whilst the paper is generally well written there are a number of grammatical errors:

Line 1 & 2: Title doesn't quite read correctly, suggest a rewording for clarity.

L31: Insert "a"after only

L33: Replace "are playing" with "play"

L34: Begin sentence "Viruses are intracellular pathogens and as such the role..."59: replace "form" with "from"

L63: Insert "was demonstrated" after REM1.2)

L85: Replace "proposed" with "propose"

L86: replace "encouraging to explore" with "which will encourage further exploration of "

L129: replace "we'll focus on" with "we concentrated on"

L131: insert "an" before ATP

L138: change "among" to "amongst"

L156: change "allows to analyse" to "allows the analysis of"

L204: Insert "showed" after previously.

L232: "root seedlings" should this be the roots of seedlings?

L235: insert "to" after "as"

L280: insert "a" after "only"

L281: change " to play" with "as playing": change CA to superscript

L307: Insert "was" after "transcription"

L320: change "display" to "displaying"

L321: change "form" to forms"

L340: "hampering" should come before viral

L365: insert"us' after "allow"

Thank you, these were corrected

Version 2

Abstract

Abstract

Plant viruses represent a risk to agricultural production and as only a few treatments exist, it is urgent to identify resistance mechanisms and factors. In plant immunity, plasma membrane (PM)-localized proteins play an essential role in sensing the extracellular threat presented by bacteria, fungi or herbivores. Viruses are intracellular pathogens and as such the role of the plant PM in detection and resistance against viruses is often overlooked. We investigated the role of the partially PM-bound Calcium-dependent protein kinase 3 (CPK3) in viral infection and we discovered that it displayed a specific ability to hamper viral propagation over CPK isoforms that are involved in immune response to extracellular pathogens. More and more evidence support that the lateral organization of PM proteins and lipids underlies signal transduction in plants. We showed here that CPK3 diffusion in the PM is reduced upon activation as well as upon viral infection and that such immobilization depended on its substrate, Remorin (REM1.2), a scaffold protein. Furthermore, we discovered that the viral infection induced a CPK3-dependent increase of REM1.2 PM diffusion. Such interdependence was also observable regarding viral propagation. This study unveils a complex relationship between a kinase and its substrate that contrasts with the commonly described co-stabilisation upon activation while it proposes a PM-based mechanism involved in decreased sensitivity to viral infection in plants.

Introduction

Viruses are intracellular pathogens, carrying minimal biological material and strictly relying on their host for replication and propagation. They represent a critical threat to both human health and food security. In particular, potexvirus epidemics like the one caused by pepino mosaic virus dramatically affect crop production1 and the lack of chemical treatments available makes it crucial to develop inventive protective methods. Unlike their animal counterparts, which enter host cells by interacting directly with the plasma membrane (PM), plant viruses have to rely either on mechanical wounding or insect vectors to cross the plant cell wall1. For this reason, only a few PM-localized proteins were identified as taking part in immunity against viruses2,3. Among them, members of the REMORIN (REM) protein family were shown to be involved in viral propagation, with varying mechanisms depending on the studied viral genera4–9. REM proteins are well-known for their heterogeneous distribution at the PM, forming nanodomains (ND), PM nanoscale environments that display a composition different from the surrounding PM10,11. Increasing evidence supports the role of ND in signal transduction, with the underlying idea that the local accumulation of proteins allows amplification and specification of the signal11. For example, Arabidopsis RHO-OF-PLANT 6 accumulates in distinct ND upon osmotic stress and auxin treatment in a dose-dependent way for the latter12. Recently, Arabidopsis REM1.2 (later named REM1.2) was demonstrated to form clusters upon exposure to the bacterial effector flg22 to support the condensation of Arabidopsis FORMIN 6 and to induce actin cytoskeleton remodeling13. However, unlike the canonical mechanism describing the accumulation of proteins in ND upon stimulation12–14, we showed previously that Solanum tuberosum REM1.3 (StREM1.3) ND were disrupted and the diffusion of individual proteins increased in response to a viral infection15. StREM1.3 lateral organization in lipid bilayers was also shown to be modified upon its phosphorylation status, both in vitro and in vivo15,16. The role of such protein dispersion upon stimuli is not understood yet and only a few similar cases are reported in the literature17,18. REM1.2 was identified as a substrate of the partially membrane-bound CALCIUM-DEPENDENT PROTEIN KINASE 3 (CPK3)19. CPK3 is involved in defense response against herbivores, bacteria and viruses and was recently proposed to be at cross-roads between pattern-triggered immunity and effector-triggered immunity15,20,21. We showed previously that transient over-expression of CPK3 in N. benthamiana was able to hamper potato virus X (PVX, potexvirus) cell-to-cell propagation15. Although partially PM localized, the role of CPK3 PM localization in immunity was never investigated for any pathogen. As the PVX cannot infect Arabidopsis thaliana, the dedicated plant model for genetic studies, we used an alternative virus model able to infect this species, the plantago asiatica mosaic virus (PlAMV)22,23 to investigate the role of CPK3 and of its PM localization in potexvirus propagation. We were able to highlight the specific role of CPK3 among other immune-related CPKs24–26 in this context. Also, we demonstrated that CPK3 membrane localization was required to hamper viral cell-to-cell propagation and, using single-particle tracking photoactivated light microscopy (spt-PALM), we discovered that CPK3 diffusion was reduced upon activation and viral infection. Interestingly, this reduction of PM diffusion depended on the expression of Group 1 REM while viral-induced REM1.2 increased PM diffusion depended on CPK3. Overall, our data allowed us to propose a model for a PM-localized mechanism involved in the reduction of potexvirus propagation, which will encourage further exploration of the involvement of the PM in viral immunity.

Results

Arabidopsis thaliana calcium-dependent protein kinase 3 (CPK3) specifically restricts PlAMV propagation

We previously showed that transient overexpression of Arabidopsis CPK3 in N. benthamiana leaves restricted the propagation of the potexvirus potato virus X (PVX)15. CPKs are encoded by a large gene family of 34 members in Arabidopsis27. To evaluate the functional redundancy between CPKs regarding viral propagation, a series of Arabidopsis lines mutated for CPK1, CPK2, CPK3, CPK5, CPK6 or CPK11, that are involved in plant resistance to various pathogens21,24–26,28, were analyzed in a viral propagation assay. Because PVX does not infect Arabidopsis, we used instead a binary plasmid encoding for the genomic structure of a GFP-tagged PlAMV22,23, a potexvirus that is capable of infecting a wide range of plant hosts, including Arabidopsis. Agrobacterium carrying PlAMV-GFP were infiltrated in A. thaliana leaves and GFP-fluorescent infection foci were observed 5 days post infiltration (dpi) (Figure 1 – figure supplement 1). The following combinations of mutants were tested: the cpk1 cpk2 double mutant25, the cpk5 cpk6 double mutant24, the triple mutant cpk5 cpk6 cpk1124 and the two quadruple mutants cpk1 cpk2 cpk5 cpk625 and cpk3 cpk5 cpk6 cpk1126. No significant difference of PlAMV-GFP infection foci area was detected between Col-0, cpk1 cpk2, cpk5 cpk6, cpk1 cpk2 cpk5 cpk6 and cpk5 cpk6 cpk11, demonstrating that CPK1, CPK2, CPK5, CPK6 and CPK11 are not involved in PlAMV propagation (Figure 1A, 1B). However, a 20 % increase of PlAMV-GFP infected area was observed in cpk3 cpk5 cpk6 cpk11 quadruple mutant relative to the control Col-0, which indicates that CPK3 could play a specific role in viral propagation. Since group 1 REMs are known substrates of CPK319, we checked REM1.2 phosphorylation specificity by the CPK isoforms tested in viral propagation (Figure 1 – figure supplement 2). CPK1, 2 and 3 displayed the strongest kinase activity on REM1.2 while CPK5, CPK6 and CPK11 displayed a residual activity. In contrast, all 6 CPKs phosphorylated the generic substrate histone, suggesting some substrate specificity for REM1.2 in vitro. Since CPK1 and CPK2 were described to be mainly localized within the peroxisome29 and endoplasmic reticulum30, respectively, we hypothesize that they likely do not interact in vivo with PM-localized REM1.231.

Arabidopsis thaliana calcium-dependent protein kinase 3 (CPK3) is specifically involved in the restriction of PlAMV cell-to-cell movement.

A. Representative images of PlAMV-GFP infection foci at 5 dpi in the different mutant backgrounds. Scale bar = 500 µm

B. Box plots of the mean area of PlAMV-GFP infection foci 5 days after infection in CPK multiple mutant lines, normalized to the mean area measured in Col-0. Three independent biological repeats were performed, with at least 47 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

C. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-1 and cpk3-2 single mutants and in CPK3 over-expressing lines (Pro35S:CPK3-HA #8.2 and Pro35S:CPK3-HA #16.2), normalized to the mean area measured in Col-0. Three independent biological repeats were performed, with at least 56 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

D. Representative images of A. thaliana plants infected with PlAMV-GFP and imaged with a CCD Camera from 10 to 17 dpi. The region of interest used for measurement of pixel intensity is circled with a white dotted line. Multicolored scale is used to enhance contrast and ranges from blue (low intensity) to red (high intensity). Scale bar = 4 cm.

E. Box plots of the mean cumulated intensity measured in infected leaves in Col-0, cpk3-2 and Pro35S:CPK3-HA #16.2 during systemic viral propagation. Two independent experiments were conducted. Statistical differences could be observed between the genotypes and time-points using a Kruskal-Wallis followed by a Dunn’s multiple comparison test (p<0.05). For clarity, only the results of the statistical test of the comparison of the different time-points (10, 14 and 17 dpi) within a genotype are displayed and are color-coded depending on the genotype.

F. Kinase activity of CPK3 dead variant. Recombinant proteins GST-CPK3 WT and K107M were incubated with REM1.21–118–6His in kinase reaction buffer in the presence of EGTA (−) or 100 µM free Ca2+ (+). Radioactivity is detected on dried gel (upper panel). The protein amount is monitored by Coomassie staining (lower panel).

G. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2 complemented lines cpk3-2/Pro35S:CPK3-myc and cpk3-2/ Pro35S:CPK3K107M-myc. Three independent biological repeats were performed, with at least 51 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001)

To further confirm a role for CPK3 in PlAMV infection, two independent knock-out (KO) lines for CPK3, cpk3-132 and cpk3-219 along with two independent lines overexpressing CPK3 (i.e., Pro35S:CPK3-HA-OE #8.2 and Pro35S:CPK3-HA-OE#16.272) (Figure 1 – figure supplement 3) were infiltrated with PlAMV-GFP. In both cpk3-1 and cpk3-2 KO lines, PlAMV-GFP propagation was significantly enhanced (40 to 60% compared with WT Col-0), whereas the over-expression lines Pro35S:CPK3-HA-OE#8.2 and Pro35S:CPK3-HA-OE#16.272 showed 10% and 20 % restriction of the foci area, respectively (Figure 1C). To assess whether CPK3 regulates viral propagation at the plant level, the propagation of PlAMV-GFP in systemic leaves was assessed in 3-week-old cpk3-2 and Pro35S:CPK3-HA-OE#16.2 lines along with Col-0 at 10, 14 and 17 days post infection (dpi) using a CCD Camera equipped with a GFP filter (Figure 1 – figure supplement 1). We observed that loss of CPK3 led to an increase of PlAMV-GFP propagation in distal leaves during the course of our assay while the overexpression of CPK3 did not hamper PlAMV-GFP to a greater extent than WT Col-0 (Figure 1D and E). This would suggest that CPK3 effect on PlAMV propagation is predominant at the site of infection. For this reason, we concentrated on foci in local leaves for further study.

CPK3 Lysine 107 functions as an ATP binding site and its substitution into methionine abolishes CPK3 activity in vitro21. In good agreement, we observed that CPK3K107M can no longer phosphorylate REM1.2 in vitro (Figure 1F). To test whether CPK3 kinase activity is required for its function during PlAMV infection, we analyzed the propagation of PlAMV-GFP in complementation lines of cpk3-2 with WT CPK3 or with CPK3K107M. While the WT CPK3 fully complemented cpk3-2 mutant, it was not the case with cpk3-2/Pro35S:CPK3K107M, which displayed larger infection foci area compared to WT Col-0 (Figure 1G).

Taken together, these results demonstrate a specific involvement for CPK3 among other previously described immune-related CPKs in limiting PlAMV infection.

PlAMV infection induces a decrease in CPK3 PM diffusion

We next wondered whether CPK3 accumulation is regulated at transcriptional and translational level upon PlAMV infection. RT-qPCR and western blots of CPK3 in Col-0 showed that both transcript and protein levels remained unchanged during PlAMV infection (Figure 2 – figure supplement 1). CPK3 is partially localized within the PM and is myristoylated at Glycine 2, a modification required for its association with membranes19. To test whether CPK3 membrane localization is required to hamper PlAMV propagation, we transformed cpk3-2 mutant with either ProUbi10:CPK3-mRFP1.2 or ProUbi10:CPK3G2A-mRFP1.2 (Figure 2A) and tested PlAMV infection. We observed that, in contrary to ProUbi10:CPK3-mRFP1.2, ProUbi10:CPK3G2A-mRFP1.2 did not complement cpk3-2 (Figure 2B). These observations indicate that CPK3 association with the PM is required for its function in inhibiting PlAMV propagation. We next analyzed the organization of CPK3 PM pool in absence or presence of PlAMV-GFP using confocal microscopy. Imaging of the surface of A. thaliana leaf epidermal cells expressing CPK3-mRFP1.2 showed that the protein displayed a heterogeneous pattern at the PM in both conditions (Figure 2C), although the limitation in lateral resolution of confocal microscopy hindered a more detailed analysis of CPK3 PM organization.

CPK3 diffusion decreases upon PlAMV infection in Arabidopsis thaliana

A. Confocal images of the secant view of A. thaliana epidermal cells of the cpk3-2/CPK3-mRFP1.2 or cpk3-2/ProUbi10:CPK3G2A-mRFP1.2. Scale bar = 5µm

B. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2 and complemented lines cpk3-2/ProUbi10:CPK3-mRFP1.2 and cpk3-2/ProUbi10:CPK3G2A-mRFP1.2. Three independent biological repeats were performed, with at least 32 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001).

C. Confocal images of the surface view of A. thaliana epidermal cells of the Col-0/ProUbi10:CPK3-mRFP1.2, infiltrated either with free GFP (“mock”) or PlAMV-GFP. Scale bar = 5µm

D. Representative trajectories of Col-0/CPK3-mEOS3.2 infiltrated either with free GFP (“mock”) or PlAMV-GFP; Scale bar = 2µm

E. Distribution of the diffusion coefficient (D), represented as log(D) for Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 8086 trajectories obtained in at least 16 cells over the course of three independent experiments.

F. Box plot of the mean peak value extracted from the Gaussian fit of log(D) distribution. Significant difference was revealed using a Mann-Whitney test. *: p<0.05.

G. Mean square displacement (MSD) over time of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Representative trajectories extracted from Figure 2D illustrate each curve. Scale bar = 1µm. Data were acquired from at least 16 cells over the course of three independent experiments.

H. Voronoi tessellation illustration of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. ND are circled in red. Scale bar = 2µm

I. Distribution of the ND diameter of Col-0/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

J. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 2I. No significant differences were revealed using a Mann-Whitney test.

K. Boxplot of the proportion of Col-0/ProUbi10:CPK3-mEOS3.2 detections found in ND five days after infiltration with free GFP (“mock”) or PlAMV-GFP. No significant differences were revealed using a Mann-Whitney test.

Thus, we used single particle tracking phospho-activated localization microscopy (sptPALM) which overcomes the diffraction limit of confocal microscopy and allows the analysis of the diffusion and organization of single molecules. We used a translational fusion of CPK3 with the true monomeric photoconvertible fluorescent protein mEOS3.233 expressed in stable transgenic Arabidopsis lines. We imaged these materials in control and upon PlAMV infection. We tracked single molecule trajectories (Figure 2D) from which CPK3 diffusion coefficient (D) was calculated. We observed that CPK3 proteins were overall mobile in control and infected conditions (log(D) > −2; Figure 2E) although CPK3 diffusion was reduced upon PlAMV infection (Figure 2F). Analysis of the mean squared displacement (MSD), describing the surface explored by single molecules overtime, showed that CPK3 displayed a more confined behavior during a PlAMV infection than in healthy conditions (Figure 2G). Additionally, we performed cluster analysis using Voronoï tessellation, a computation method that segments super-resolution images into polygons based on the local molecule density34. Voronoï analysis showed that no difference occurred in CPK3 cluster size or proportion of protein localized in cluster upon viral infection (Figure 2H-K). Taken together, these results show that CPK3 diffusion parameters were modified upon PlAMV infection, although the nano-organization of the proteins was maintained.

Truncation of CPK3 auto-inhibitory domain induces its confinement and accumulation in PM ND

CPK3 bears an auto-inhibitory domain that folds over the kinase domain and inhibits its kinase activity in the absence of calcium35,36. The truncation of this domain along with the C-terminal regulatory domain results in a calcium-independent, constitutively active CPK3 (CPK3CA)24 that is lethal when stably expressed in Arabidopsis37. For this reason, CPK3CA was transiently expressed in N. benthamiana for further analysis. We observed that although both ProUbi10:CPK3-mRFP1.2 and ProUbi10:CPK3CA-mRFP1.2 were partially cytosolic when transiently expressed in N. benthamiana (Figure 3 – figure supplement 1), ProUbi10:CPK3CA-mRFP1.2 displayed a PM organization in domains discernable by confocal microscopy (Figure 3A).

PlAMV-induced activation of CPK3 in N. benthamiana induces its confinement and clustering in PM domains

A. Confocal images of the surface view of N. benthamiana epidermal cells transiently expressing ProUbi10:CPK3-mRFP1.2, ProUbi10:CPK3-mRFP1.2 + PlAMV-GFP or ProUbi10:CPK3CA-mRFP1.2. Scale bar = 5µm

B. Representative trajectories of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3.2-mEOS3.2 + PlAMV and ProUbi10:CPK3CA-mEOS3.2. Scale bar = 2µm.

C. Distribution of the diffusion coefficient (D), represented as log(D) for ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. Data were acquired from at least 6144 trajectories obtained in at least 15 cells over the course of three independent experiments.

D. Box plots of the fraction of immobile proteins (log(D)<-2). Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.005).

E. Mean square displacement (MSD) over time of fast (circle) or slow (triangle) diffusing ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP, ProUbi10:CPK3CA-mEOS3.2. Standard error is displayed from three independent experiments.

F. Voronoï tessellation illustration of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. ND are circled in red. Scale bar = 2µm

G. Distribution of the ND diameter of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2

H. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 3G. No significant differences were revealed using a Kruskal-Wallis followed by a Dunn’s multiple comparison test.

I. Boxplot of the proportion of detections found in ND of ProUbi10:CPK3-mEOS3.2, ProUbi10:CPK3-mEOS3.2 + PlAMV-GFP and ProUbi10:CPK3CA-mEOS3.2. Significant differences were revealed using a Kruskal-Wallis followed by a Dunn’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.005).

J. Left: Confocal images of the surface view of N. benthamiana epidermal cells transiently expressing ProUbi10:CPK3CA-mRFP1.2 and infiltrated with either DMSO or 50 µg/mL fenpropimorph. Scale bar = 5µm; Right: Box plot of the mean Spatial Clustering Index (SCI) of CPK3CA. At least three experiments were performed, with at least 10 cells per experiment; statistical significance was determined using a Student t-test, ****: p < 0.0001.

K. Left: Confocal images of the surface view of N. benthamiana epidermal cells transiently co-expressing ProUbi10:CPK3CA-mRFP1.2 with active or dead SAC1, mutated for its phosphatase activity. Scale bar = 5µm; Right: Box plot of the mean SCI of CPK3CA. At least three experiments were performed, with at least 10 cells per experiment; statistical significance was determined using a Student t-test, ****: p < 0.0001.

We analyzed the dynamics of ProUbi10:CPK3-mEOS3.2 and ProUbi10:CPK3CA-mEOS3.2 by spt-PALM (Figure 3B). We observed that the fraction of immobile ProUbi10:CPK3CA-mEOS3.2 molecules (log(D) < −2) is more abundant than for ProUbi10:CPK3-mEOS3.2 in absence of viral infection while no significant difference could be deciphered upon PlAMV infection (Figure 3C and D). In addition, MSD analysis showed that the motion of CPK3-mEOS3.2 mobile fraction is less confined than the one of CPK3CA-mEOS3.2 and CPK3-mEOS3.2 upon PlAMV infection (Figure 3E). Overall, this indicates that the mobile and immobile fraction of CPK3 is affected upon PlAMV infection and upon truncation of its auto-inhibitory domain, respectively.

Cluster analysis of CPK3 was performed using tessellation on the localization data obtained with spt-PALM (Figure 3F). Although we did not observe any significant differences in distribution of cluster sizes between all compared conditions(Figure 3G and H), CPK3CA displayed a significantly higher proportion of proteins localized in ND compared to CPK3 (Figure 3I).

While the mechanism(s) governing the clustering of membrane proteins are not fully described, it is widely accepted that lateral organization involves – to some extent – protein-lipid interactions and lipid-lipid organization10,11,38. We observed that the integrity of ProUbi10:CPK3CA-mRFP1.2 organization relied on sterols and phosphoinositides. Indeed, treatment with fenpropimorph, a well-described inhibitor of sterol biosynthesis39, abolished CPK3CA ND organization (Figure 3J), and the co-expression of ProUbi10:CPK3CA-mRFP1.2 with the yeast phosphatidylinositol-4-phosphate (PI4P)-specific phosphatase SAC1 targeted to the PM40, led to sparser and bigger domains (Figure 3K), which suggested that PI4P is not required for CPK3CA ND formation but for its regulation.

All together these observations show that either removal of the auto-inhibitory domain or infection with PlAMV-GFP modifies CPK3 dynamics within the PM ; a mechanism which could be at stake for kinase activation.

PlAMV infection induces an increase in REM1.2 PM diffusion

Group 1 REMs are one of the targets of CPK3 and we previously demonstrated that the restriction of PVX propagation by CPK3 overexpression depended on endogenous group 1 NbREMs15. Four REM isoforms belong to the group 1 in Arabidopsis: REM1.1, REM1.2, REM1.3 and REM1.441. REM1.2 and REM1.3 are amongst the 10% most abundant transcripts in Arabidopsis leaves while REM1.1 was not detected in recently published leaf transcriptomes and proteomes42. Therefore, we focused on the three isoforms REM1.2, REM1.3 and REM1.4. REMs are described as scaffold proteins43, for which physiological function depends on the proteins they interact with and their phosphorylation status44. As recently described in45, REM1.2 and REM1.3 share 95% of their interactome, suggesting that they are functionally redundant. To address this, we isolated single T-DNA mutants rem1.2, rem1.3 and rem1.4 (SALK_117637.50.50.x, SALK_117448.53.95.x and SALK_073841.47.35, respectively) and crossed them to obtain the double mutant rem1.2 rem1.3 and the triple mutant rem1.2 rem1.3 rem1.4 (Figure 4 – figure supplement 1). We did not notice any obvious defects in the growth and development of seedlings and adult plants for the single, double and triple mutants, when grown under our conditions (Figure 4 – figure supplement 2). No difference in PlAMV-GFP propagation could be observed in the single mutants, compared to Col-0 (Figure 4A). However, the double mutant rem1.2 rem1.3 showed a significant increase of infection foci area compared to Col-0, which was further enhanced in rem1.2 rem1.3 rem1.4 triple KO mutant. Such additive effect of multiple mutations shows that REM1.2, REM1.3 and REM1.4 are functionally redundant regarding PlAMV cell-to-cell propagation. Finally, PlAMV-GFP systemic propagation was followed in whole plants every 3-4 days from 10 to 17 dpi and rem1.2 rem1.3 rem1.4 displayed an increased infection surface of systemic leaves compared to Col-0 (Figure 4B and C), suggesting that group 1 REMs are involved in both local and systemic propagation of PlAMV-GFP.

Group 1 REM hampers PlAMV-GFP cell-to-cell propagation and REM1.2 diffusion increases upon infection

A. Box plots of the mean area of PlAMV-GFP infection foci in rem1.2, rem1.3, rem1.4 single mutants along with rem1.2 rem1.3 double mutant and rem1.2 rem1.3 rem1.4 triple mutant. Three independent biological repeats were performed, with at least 36 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.05).

B. Box plots of the mean cumulated intensity measured in infected leaves in Col-0 and rem1.2 rem1.3 rem1.4 during systemic viral propagation. Two independent experiments were conducted. Statistical significance of the difference between Col-0 and rem1.2 rem1.3 rem1.4 at each time point was assessed using a Mann-Whitney test. *: p<0.05, **: p<0.01

C. Representative images of A. thaliana plants infected with PlAMV-GFP and imaged with a CCD Camera from 10 to 17 dpi. Systemic leaves are circled with a white dotted line. Multicolored scale is used to enhance contrast and ranges from blue (low intensity) to red (high intensity). Scale bar = 4 cm.

D. Confocal images of the surface view of A. thaliana epidermal cells of the Col-0/ProUbi10:mRFP1.2-REM1.2, infiltrated either with free GFP (“mock”) or PlAMV-GFP. Scale bar = 5µm

E. Representative trajectories of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm

F. Distribution of the diffusion coefficient (D), represented as log(D) for Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 28638 trajectories obtained from at least 16 cells over the course of three independent experiments.

G. Box plot of the mean peak value extracted from the Gaussian fit of log(D) distribution. Significant difference was revealed using a Mann-Whitney test. *: p<0.05.

H. Mean square displacement (MSD) over time of Col-0/ProUbi10:mEOS3.2-REM1.2 infiltrated either with free GFP (“mock”) or PlAMV-GFP. Representative trajectories extracted from Figure 4E illustrate each curve. Scale bar=1µm.

I. Voronoi tessellation illustration of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. ND are circled in red. Scale bar = 2µm

J. Distribution of the ND diameter of Col-0/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

K. Box plot representing the mean peak value of ND diameter extracted from the Gaussian fit of Figure 4J. No significant difference was revealed using a Mann-Whitney test.

L. Boxplot of the proportion of Col-0/ProUbi10:mEOS3.2-REM1.2 detections found in ND five days after infiltration with free GFP (“mock”) or PlAMV-GFP. No significant difference was revealed using a Mann-Whitney test.

Given their role in cell-to-cell viral movement, we checked whether group 1 REM expression was modified upon infection. RT-qPCR and western blots showed that neither transcripts nor protein levels were modified upon PlAMV infection (Figure 4 – figure supplement 3). Since REM1.2 and REM1.3 share a large part of their interactome and show functional redundancy regarding PlAMV infection, we decided to focus on REM1.2 for further investigations. Confocal imaging of the surface of epidermal cells of Col-0/ProUbi10:mRFP1.2-REM1.2 showed a rather heterogeneous distribution at the PM, although less striking than previously described when observed in root31 (Figure 4D). REM1.2 was next fused to mEOS3.2 and stably expressed in Col-0 to conduct spt-PALM (Figure 4E). The diffusion coefficient of ProUbi10:mEOS3.2-REM1.2 was significantly increased upon PlAMV infection (Figure 4F and G). Moreover, its MSD was increased to a similar extent as what was previously observed with StREM1.3 during a PVX infection15 (Figure 4H), although REM1.2 is overall more mobile than StREM1.3. Tessellation analysis of protein localization did not show any difference in ND organization, whether in size or regarding the enrichment of proteins in ND (Figure 4I-L).

Taken together, these results show that group 1 REMs are functionally redundant regarding their ability to hamper PlAMV propagation. PlAMV infection promoted an increased diffusion of REM1.2, in the same way as PVX did with StREM1.3. The conservation of such mechanism between plants of different families is indicative of its physiological importance.

PlAMV-induced changes in REM1.2 and CPK3 plasma membrane dynamics are interdependent

Group 1 REMs from A. thaliana were previously identified as in vitro substrates of CPK319. Moreover, untargeted immunoprecipitation experiments coupled to mass spectrometry identified CPK3 as an interactor of REM1.2 in A. thaliana45. We wanted to assess the functional link between CPK3 and group 1 REM in potexvirus propagation by knocking out CPK3 into the rem1.2 rem1.3 rem1.4 mutant background. We isolated two independent CRISPR-generated rem1.2 rem1.3 rem1.4 cpk3 #1 and rem1.2 rem1.3 rem1.4 cpk3 #2 quadruple mutants (Figure 5 – figure supplement 1). We did not observe any developmental defect in these lines when grown under controlled conditions (Figure 5 – figure supplement 2). The analysis of PlAMV-GFP propagation showed that no significant additive effect could be observed between the quadruple mutant lines, cpk3-2 and rem1.2 rem1.3 rem1.4 (Figure 5A and 5B). This indicates that group 1 REMs and CPK3 function in the same signaling pathway to inhibit PlAMV propagation.

Group 1 REMs and CPK3 are in the same functional pathway and regulate each other PM diffusion upon PlAMV infection

A. Box plots of the mean area of PlAMV-GFP infection foci in cpk3-2, rem1.2 rem1.3 rem1.4 triple mutant and rem1.2 rem1.3 rem1.4/cpk3 #1 and #2 quadruple mutants. Three independent biological repeats were performed, with at least 23 foci per experiment and per genotype. Significant differences were revealed using a One-Way ANOVA followed by a Tukey’s multiple comparison test. Letters are used to discriminate between statistically different conditions (p<0.0001).

B. Representative images of PlAMV-GFP infection foci at 5 dpi in the different mutant backgrounds. Scale bar = 500µm.

C. Representative trajectories of cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with either free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm.

D. Distribution of the diffusion coefficient (D), represented as log(D) for cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with either free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 20462 trajectories obtained from at least 11 cells over the course of three independent experiments.

E. Box plot representing the mean peak value extracted from the Gaussian fit of the distribution of the diffusion coefficient (D) represented in Figure 5D. No significant difference was revealed using a Mann-Whitney test.

F. Mean square displacement (MSD) over time of cpk3-2/ProUbi10:mEOS3.2-REM1.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

G. Representative trajectories of rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Scale bar = 2µm.

H. Distribution of the diffusion coefficient (D), represented as log(D) for rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP. Data were acquired from at least 11724 trajectories obtained from at least 10 cells over the course of three independent experiments.

I. Box plot representing the mean peak value extracted from the Gaussian fit of the distribution of the diffusion coefficient (D) represented in Figure 5H. No significant difference was revealed using a Mann-Whitney test.

J. Mean square displacement (MSD) over time of rem1.2 rem1.3 rem1.4/ProUbi10:CPK3-mEOS3.2 five days after infiltration with free GFP (“mock”) or PlAMV-GFP.

We wanted to know whether the increased diffusion of REM1.2 observed on PlAMV infection was dependent on CPK3. Using spt-PALM, we obtained the diffusion parameters of ProUbi10:REM1.2-mEOS3.2 expressed in the cpk3-2 mutant background. Strikingly, we observed that both the diffusion coefficient and MSD of REM1.2 were not anymore affected during a viral infection (Figure 5C-F), showing that REM1.2 PM lateral diffusion upon PlAMV infection depends on CPK3. In a similar manner as in Col-0, REM1.2 clustering upon PlAMV infection in cpk3-2 background did not display any difference to the mock-infected plants (Figure 5 – figure supplement 3).

Moreover, we wondered whether the reciprocal effect was true for the diffusion of CPK3 in the absence of group 1 REMs. Similarly, the diffusion coefficient and the MSD of mEOS3.2-CPK3 in rem1.2 rem1.3 rem1.4 triple KO background remained the same during an infection compared to control condition (Figure 5G-J), unlike what was observed in a Col-0 background (Figure 2D-G). This result indicated that the confinement of CPK3 proteins upon viral infection depended on the presence of group 1 REMs. Moreover, contrarily to the Col-0 background, the rem1.2 rem1.3 rem1.4 displayed reduced protein concentration upon PlAMV infection (Figure 5 – figure supplement 3). This showed that group 1 REMs might play a role in CPK3 domain regulation upon viral infection.

As CPK3CA was shown to inhibit potexvirus propagation more efficiently than the full-length protein15, we tested CPK3CA ability to alter REM diffusion (Figure 5 – figure supplement 4). Upon transient co-expression with CPK3CA, REM diffusion was significantly increased. As shown in Figure 2B, CPK membrane anchor is crucial for its role in viral propagation. We tested if CPK3CA-G2A was still able to hinder REM diffusion, which was not the case. All these data support a specific role of PM-bound CPK3 in REM increased diffusion upon viral infection.

We then investigated whether CPK3 and REM would colocalize in absence or presence of the virus. Using confocal microscopy, we showed that they randomly colocalized in both situations (Figure 5 – figure supplement 5), the interaction between the kinase and its substrate probably occurring in a narrow spatiotemporal window.

Taken all together, those results show a strong inter-dependence of group 1 REMs and CPK3 both in their physiological function and in their PM lateral diffusion upon PlAMV infection.

Discussion

CPK3 specific role in viral immunity is supported by its PM organization

Although calcium-mediated signaling is suspected to be involved in viral immunity, only few calcium-modulated proteins are described as playing a role in viral propagation15,46. We showed here the crucial role of CPK3 over other immunity-related CPK isoforms24,25 by a reverse genetic approach in Arabidopsis. The precise role of CPK3 in viral immunity remains to be determined. CPK3 phosphorylates actin depolymerization factors to modulate the actin cytoskeleton21,47, a key player in host-pathogen interaction48. In particular, potexviruses induce the remodeling of the actin cytoskeleton to organize the key steps of their cycle, whether it is replication, intra-cellular movement or cell-to-cell propagation49,50. PlAMV replication and/or movement could be affected by CPK3-mediated alteration of the cytoskeleton mesh.

The specificity of a calcium-dependent kinase in a given biological process is determined by its expression pattern, subcellular localization and substrate specificity27 Controlled subcellular localization ensures proximity with either stimulus or substrate and here we showed that, similarly to Arabidopsis CPK652 and Solanum tuberosum CPK553, the disruption of CPK3 membrane anchor led to a loss-of-function phenotype (Figure 2B). Interestingly, we observed a reduction in CPK3 PM diffusion upon PlAMV infection, suggesting that not only membrane localization but also protein organization at the PM is important during viral immunity. Moreover, PlAMV-induced CPK3 PM confinement was reminiscent of the diffusion parameters displayed by CPK3CA, hinting that viral infection, kinase activation and lateral diffusion are linked. However, it is necessary to remain careful as a truncated protein deprived of its auto-inhibitory domain does not reflect the controlled and context-dependent activation of CPK3. Indeed, stable expression of CPK3CA was previously shown to be lethal37. CPK3CA ever-activated state might lead to stable or unspecific interaction with protein partners along with erratic phosphorylation of substrates, which could explain the PM domains formed by CPK3CA at the PM.

CPK3 nanoscale dynamics upon viral infection might offer another layer of specificity to convey the appropriate response to a given stimulus by ensuring proximity with specific regulators or substrates. It would be interesting to explore whether the reduction of CPK3 diffusion observed upon PlAMV infection is specific to this virus or if it can be extended to other viral species, genera and even pathogens. Indeed, it was recently shown that CPK3 transcription was enhanced upon infection by viruses from different genera54. Moreover, CPK3 regulates herbivore responses by phosphorylating transcription factors20, is activated by flg22 in protoplasts19 and is proposed to be the target of a bacterial effector to disrupt immune response21. Finally, the diffusion and clustering of other PM-localized CPKs could be investigated as no experimental data exist yet regarding their PM nano-organization. It would be especially relevant to describe these parameters for the CPK isoforms phosphorylating ND-organized NADPH oxidases14,25,28.

Potential functions for REM1.2 increased diffusion upon PlAMV

REM proteins display a wide range of physiological functions and are proposed to function as scaffold proteins43,44,55. Group 1 REMs have been shown to be involved in viral immunity, playing apparent contradictory roles depending on the virus genera8,44,56. Herein, we show through a combination of genetic and biochemical analysis that the three most expressed isoforms of group 1 Arabidopsis REMs were functionally redundant in inhibiting PlAMV propagation (Figure 4A). REMs are anchored at the PM through their C-terminal sequence44,56–58 but despite displaying a similar PM attachment, potato StREM1.3 is static and forms well defined membrane compartments57 while Arabidopsis REM1.2 appeared mobile in leaves, with small and potentially short-lived membrane domains of around 70nm (Figure 4F-J). Beyond clear organizational differences, both REM1.2 and StREM1.3 showed an increased mobility upon viral infection in N. benthamiana and Arabidopsis15 (Figure 4E-H), which is at contrast with the canonical model linking protein activation with its stabilization into ND11,59. Particularly, REM1.2 was recently shown to form ND upon elicitation by a bacterial effector13 or upon exposition to bacterial membrane structures60. Conservation across plant and virus species of such increased PM diffusion of group 1 REMs indicates that this specific mechanism is crucial for plant response to potexviruses, although its role remains to be deciphered. It was suggested, in the context of tobacco rattle virus infection, that REM1.2 clusters led to an increased lipid order of the PM and a morphological modification of plasmodesmata (PD), inducing a decrease of PD permeability31. As REM1.2 does not accumulate at PD upon PlAMV infection (Figure 4 – figure supplement 4), it might indirectly influence PD permeability through mechanisms yet to be discovered. We previously showed that StREM1.3 modulated the formation of lipid phases in vitro61 while Medicago truncatula SYMREM1 was recently shown to stabilize membrane topology changes in protoplasts61. Therefore, REM1.2 increased diffusion upon viral infection might alter lipid organization, with consequences on PM and PD-localized proteins62. Based on REM’s putative scaffolding role43,63, its increased PM diffusion might modify the stability of REM-supported complexes and induce subsequent signaling, similarly to the way Oryza sativa REM4.1 orchestrates the balance between abscisic acid and brassinosteroid pathways by interacting with different protein kinases64.

A PM-localized mechanism involved in hampering viral propagation

Increasing evidence supports the role of PM proteins in plant antiviral mechanism65. However, while the nano-organization of PM proteins involved in bacterial or fungal immunity begins to be addressed13,60,66,67, there is still only scarce information in a viral infection context. Our observations pointed towards an interdependence of REM and CPK3 in their PM diffusion upon PlAMV infection. In particular, CPK3-dependent REM increased diffusion in a potexvirus infection is consistent with our previously reported PM diffusion increase of StREM1.3 phosphomimetic mutant, reminiscent of StREM1.3 behavior upon viral infection15. Moreover, we recently published that in vitro CPK3-phosphorylated StREM1.3 presented disrupted domains on model membranes compared to the non-phosphorylated protein16. The data presented in this paper further support the predominant role of CPK3 in viral-induced REM1.2 diffusion (Figure 5C-F). Since the actin cytoskeleton was shown to favor nanometric scale ND including those of group 1 REM68, CPK3 mediated regulation of the actin cytoskeleton could regulate REM1.2 PM nanoscale organization. We also discovered the essential role of group 1 REMs in the lateral organization of CPK3 upon viral infection (Figure 5G-J), which is further supported by the difference between CPK3 clustering parameters when expressed in N. benthamiana or in Arabidopsis: CPK3 was enriched in ND upon PlAMV infection when transiently expressed in N. benthamiana but not in Arabidopsis (Figure 2H, 2K, 3F and 3I). The discrepancy between both species could be linked to N. benthamiana REM1 ability to form stable ND, discernable by confocal microscopy5 while we observed small and unstable domains of REM1.2 in Arabidopsis leaves (Figure 4I-K). Moreover, CPK3 ND organization was disrupted in rem1.2 rem1.3 rem1.4 triple KO background upon PlAMV infection (Figure 5 – figure supplement 3H), indicating that group1 REMs are crucial for the spatial organization of CPK3 during a viral infection. Furthermore, the sensitivity of CPK3CA domains to the same lipid inhibitors as StREM1.357 (Figure 3J,K) hints that REM1.2 lipid environment might be required for CPK3 nanoscale organization.

Although the data obtained here do not allow us to determine the sequence of events orchestrating REM1.2 and CPK3 dynamic interaction, hypotheses can be formulated (Figure 6). REM and CPK3 interact in the absence of any stimulus15,45 but as they display drastically different diffusion parameters (Figure 2E, Figure 4F), likely only a small part of proteins interact at basal state. PlAMV infection induces a REM-dependent confinement of CPK3 (Figure 2 and Figure 5). Whether CPK3 confinement is concomitant to its activation remains to be confirmed. Such modification of CPK3 mobility might result in an increase probability of interaction with REM1.2, which might lead to increased phosphorylation events and a subsequent increase in mobility of REM1.2. In addition to the above-mentioned putative roles of REM’s increased diffusion, such “kiss-and-go” mechanism between a kinase and its substrate might also be considered as a negative feedback loop to hamper constitutive activation of the system. It was recently shown that the PM organization of a receptor-like kinase and its co-receptor relied on the expression of a scaffold protein67,69, REM might go beyond the role of substrate and be essential to maintain the required PM environment of its cognate kinase to ensure proper signal transduction. Although the complexity of potexvirus’ lifecycle as well as technical limitations impair a thorough challenge of this hypothesis, our data support the significance of fine PM compartmentalization and spatio-temporal dynamics in signal transduction upon viral infection in plants while it explores new concepts underlying plant kinases PM organization.

Proposed “kiss-and-run” model describing REM1.2 and CPK3 interdependence

In healthy conditions, REM1.2 displays a smaller mobility at the PM than CPK3. Random interactions between the two proteins can occur.

Upon PlAMV infection, CPK3 shows a REM-dependent confined behavior which might be concomitant to its activation. Simultaneously, a PlAMV infection leads to a CPK3-dependent increased diffusion of REM1.2, which might be a consequence of its phosphorylation by CPK3.